---
output: cynkratemplate::readme_document
---

<!-- README.md and index.md are generated from README.Rmd.
     Edit that file and render it the usual way: rmarkdown::render(),
     devtools::build_readme(), or the Knit button. The cynkratemplate
     package must be installed; it supplies the output format. -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
pkgload::load_all()
```

# winch

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html)
[![R build status](https://github.com/r-prof/winch/workflows/rcc/badge.svg)](https://github.com/r-prof/winch/actions)
[![CRAN status](https://www.r-pkg.org/badges/version/winch)](https://CRAN.R-project.org/package=winch)
<!-- badges: end -->

Winch provides stack traces for call chains that cross between R and C function calls.
This is a useful tool for developers of R packages where a substantial portion of the code is C or C++.


## Installation

Install the released version of winch from [CRAN](https://cran.r-project.org/) with:

``` r
install.packages("winch")
```

Install the development version from GitHub with:

``` r
# install.packages("pak")
pak::pak("r-prof/winch")
```



## Example

Below is an example where an R function calls into C which calls back into R. 
Note the second-to-last entry in the trace:

```{r example}
library(winch)

foo <- function() {
  winch_call(bar)
}

bar <- function() {
  winch_trace_back()
}

trace <- foo()

# Keep the render reproducible: the instruction pointers are randomised by
# ASLR, and the shared object is loaded from a temporary directory whose
# name is new on every load -- and the varying path width re-aligns the
# whole table. Show the columns that are stable.
trace$ip <- NULL
trace$pathname <- basename(trace$pathname)
# A frame whose symbol the linker cannot resolve falls back to its address,
# which is randomised too.
trace$func <- sub("^[0-9a-f]{8,}$", "<unresolved>", trace$func)
trace
```

`rlang::entrace()` checks if winch is installed, and adds a native backtrace.
As this cannot be easily demonstrated in a knitr document, the output is copied from a GitHub Actions run.

```{r error, eval = FALSE}
options(
  error = rlang::entrace,
  rlang_backtrace_on_error = "full",
  rlang_trace_use_winch = TRUE
)

vctrs::vec_as_location(quote, 2)
```

```
Error: Must subset elements with a valid subscript vector.
✖ Subscript has the wrong type `function`.
ℹ It must be logical, numeric, or character.
Backtrace:
    █
 1. └─vctrs::vec_as_location(quote, 2)
 2.   └─`/vctrs.so`::vctrs_as_location()
 3.     └─`/vctrs.so`::vec_as_location_opts()
 ```


## How does it work?

winch uses a very simple heuristic.
R's traceback (and also profiling) infrastructure introduces the notion of a "context".
Every call to an R function opens a new context and closes it when execution of the function ends.
Unfortunately, no new context is established for native code called with `.Call()` or `.External()`.
Establishing contexts expends precious run time, so this may be the reason for the omission.

To work around this limitation, the source code of all R functions along the call chain is scanned for instances of `.Call` and `.External`.
The native call stack (obtained via [libunwind](https://github.com/libunwind/libunwind) or [libbacktrace](https://github.com/ianlancetaylor/libbacktrace)) is scanned for chunks of code outside of `libR.so` (R's main library) --
these are assumed to correspond to `.Call()` or `.External()`.
The native traces are embedded as artificial calls into the R stack trace.

## Limitations

- The matching will not be perfect, but it may still lead to faster discovery of the cause of an error.
- On Windows winch only works on x64, and there the traces can be obtained only for one shared library at a time. 
See `winch_init_library()` for details.

---

## Code of Conduct

Please note that the winch project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/0/CODE_OF_CONDUCT.html).
By contributing to this project, you agree to abide by its terms.
