---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
pkgload::load_all()
```

# winch

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
<!-- badges: end -->

The goal of winch is to provide stack traces that combine R and C function calls.
This is primarily useful for developers of R packages where a substantial portion of the code is C or C++.


## Installation

Once on CRAN, you can install the released version of winch from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("winch")
```

Install the development version from GitHub with:

``` r
devtools::install_github("r-lib/winch")
```

## Example

This is an example where an R function calls into C which calls back into R, see the second-to-last entry in the trace:

```{r example}
library(winch)

foo <- function() {
  winch_call(bar)
}

bar <- function() {
  winch_trace_back()
}

foo()
```

`winch_entrace()` is a drop-in replacement for `rlang::entrace()`.
This cannot be easily demonstrated in a knitr document, see [this GitHub Actions run](https://github.com/r-prof/winch/runs/895443204?check_suite_focus=true#step:11:53) for an example.

```{r error, eval = FALSE}
library(winch)

options(error = winch_entrace, rlang_backtrace_on_error = "full")

foo <- function() {
  winch_call(function() bar())
}

bar <- function() {
  stop("oops")
}

foo()
```

## How does it work?

It's a very crude heuristic.
R's traceback (and also profiling) infastructure introduces the notion of a "context".
Every call to an R function opens a new context, and closes it when execution of the function ends.
Unfortunately, no new context is established for native code called with `.Call()` or `.External()`.
Establishing contexts requires precious run time, this might be the reason for this omission.

To work around this limitation, the source code of all R functions along the call chain is scanned for instances of `.Call` and `.External`.
The native call stack (obtained via libc's [`backtrace_symbols()`](https://www.gnu.org/software/libc/manual/html_node/Backtraces.html) is scanned for chunks of code outside of `libR.so` (R's main library) -- these are assumed to correspond to `.Call()` or `.External()`.
The native traces are embedded as artificial calls into the R stack trace.

## Limitations

- The matching will not be perfect, it still may lead to quicker discovery of the cause of an error.
- Currently Linux and macOS only.
