---
title: "Joint backtrace"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Joint backtrace}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(winch)
```


- Intro: what is this document about, outline

    - fresh start: more controlled solution to the complicated problem of combining stack traces

## Backtraces

- What is it? How to generate in R: `sys.calls()`

- What about native code: Initial SO question: "Can an application programmatically generate its own stacktrace?" https://stackoverflow.com/q/8031742/946850

    - cite reasons for unwinding from https://news.ycombinator.com/item?id=11477039
    
    - provided by the compiler or by specialized libraries

    - main result: sequence of IP (instruction pointers); function name, file+line and even library name must be deduced manually; often with library support

- for now we're considering static code; link above links to the problem of dynamic unwinding, *portable* static unwinding is a mess already

- R always has been an interface language, good integration with calling into foreign interfaces. R code often calls into specialized native code for speed or to avoid reimplementation in R.

- task: when R code calls into native code (which then might call back into R), generate backtrace that represents chain of function calls from both domains

    - current state: native code hidden from R backtrace
    
    - current state: R code is represented as sequence of function calls in `libR.so`
    
    - solution: inject relevant bits of native backtrace into R backtrace

- demo: `winch_call()` calls an R function from C code, solely for demonstration. The R function is given as argument. The roundtrip into C code is shown as `.Call()` entries without further detail.

```{r}
library(winch)

foo <- function() {
  winch_call(bar)
}

bar <- function() {
  winch_call(baz)
}

baz <- function() {
  sys.calls()
}

foo()
```

- `winch_trace_back()` returns the native backtrace as an R data frame. When reviewing the first entries, the native calls from `winch_call()` are visible.


```{r}
baz <- function() {
  winch_trace_back()
}

head(foo(), 25)
```

- Relevant information: native function name, library name

- more details to add: file+line, remains open issue

- `winch_trace_back()` is a low-level function, corner stone for joint backtraces provided by `winch_add_trace_back()`

## rlang integration

- must collect backtrace when error occurs, in `trace_back()`

- https://github.com/r-lib/rlang/pull/1039

- demo:

```{r}
options(rlang_trace_use_winch = TRUE)

baz <- function() {
  rlang::trace_back()
}

foo()
```


- demo: The vctrs package uses C for performance-critical parts, and often calls back into R for e.g. S3 dispatch or error reporting. Example: backtrace after an error.

```{r error, eval = FALSE}
options(
  error = rlang::entrace,
  rlang_backtrace_on_error = "full",
  rlang_trace_use_winch = 1L
)

vctrs::vec_as_location(quote, 2)
```

```
Error: Must subset elements with a valid subscript vector.
✖ Subscript has the wrong type `function`.
ℹ It must be logical, numeric, or character.
Backtrace:
    █
 1. ├─vctrs::vec_as_location(quote, 2)
 2. │ └─`/vctrs.so`::vctrs_as_location()
 3. └─rlang::cnd_signal(...) /home/kirill/git/R/vctrs/R/subscript-loc.R:64:2
```

- The remainder of the document discusses technical details and caveats.

## Retrieving IP

- several options, unfortunately no "one size fits all" solution

- solutions that retrieve IP also find function name (and sometimes file+line) out of the box

- module name separately

- tried execinfo -- pretty bad, not on Windows

- tried libunwind -- great, not on Windows: https://libunwind-devel.nongnu.narkive.com/ahavTcFa/libunwind-and-windows

    - two versions: http://lists.llvm.org/pipermail/cfe-dev/2016-September/050650.html

    - linux version: nongnu-libunwind,  https://www.nongnu.org/libunwind/, https://github.com/libunwind/libunwind

     - clang version: https://bcain-llvm.readthedocs.io/projects/libunwind/en/latest/, https://github.com/llvm/llvm-project/blob/368c02e3ec44e5418626f46abebcc22a69c7f66d/libunwind/include/libunwind.h

- tried libbacktrace -- slightly worse than libunwind, works on Windows mingw64 with a patch, includes file paths, crashes on clang

- reviewed boost stacktrace -- only for printing, https://www.boost.org/doc/libs/master/boost/stacktrace/detail/collect_unwind.ipp

- stacktrace, from 2009: http://stacktrace.sourceforge.net/

- fully native route not considered, e.g. https://programmer.help/blogs/using-the-_unwind_backtrace-function-to-grab-the-backtrace-of-the-c-c-stack.html, http://www.nynaeve.net/?p=99

- need both libunwind (for clang) and libstacktrace (for Windows)

### libunwind gotchas

- need library on Linux, on macOS `-lSystem` seems sufficient

- no file+line information

### libbacktrace gotchas

- vendored library, shipped as part of GCC and also available standalone, with tests!

- libbacktrace btest works out of the box in mingw32, sjlj

- libbacktrace works on mingw64 with gcc patch: https://github.com/r-windows/rtools-packages/pull/148, https://github.com/gcc-mirror/gcc/pull/48

    - documented in libbacktrace issue: https://github.com/ianlancetaylor/libbacktrace/issues/43#issuecomment-687858320

    - gcc bug reported: https://gcc.gnu.org/bugzilla/show_bug.cgi?id=96948

    - patch made available independently

    - patch submitted to rtools-packages: https://github.com/r-windows/rtools-packages/pull/148

    - works somewhat if using region addresses, can slightly patch libbacktrace to achieve this with stock gcc: https://github.com/ianlancetaylor/libbacktrace/pull/54, rejected upstream, good enough for us

- almost as good as libunwind when looking up symbol information, observed difference in some `libR.so` symbols

    ! https://github.com/ianlancetaylor/libbacktrace/pull/38 seems irrelevant

- passes tests on macOS, crashes in winch package

## Retrieving module name

- no portable solution

- existing code from gperftools project packaged as separate package procmaps, on CRAN

    - too complex, update submitted that gets rid of `./configure` script and more unneeded code.

    - rewrite as simple C code, perhaps include into libbacktrace: https://github.com/ianlancetaylor/libbacktrace/issues/53
    
## Retrieving file + line

- built into libbacktrace

    - doesn't work on macOS

- addr2line equivalent on Mac? https://sourceforge.net/p/elftoolchain/code/HEAD/tree/trunk/addr2line/addr2line.c, https://stackoverflow.com/q/11556321/946850


## Tests

- test on Windows with correct gcc and with wrong gcc, and with old rtools, and with i386

    ! path of main executable via whereami, is it helpful at all? Branch f-windows-whereami

    ! Build with debug info using R CMD INSTALL --debug . or by setting the DEBUG environment variable, adds -gdwarf-2 -- this is wonderful

    ! for pecoff, need to add symbolic and debug information for every library, not only for the main one: https://github.com/ianlancetaylor/libbacktrace/issues/53

        ! workaround: add symbolic and debug information for active file

    ! use instruction pointer in absence of syminfo (both libunwind and libbacktrace)

    ! i386 won't work, backtrace is terminated prematurely

        ! exclude in tests

! we can build libbacktrace on macOS with clang, but doesn't run

    - btest gives "libbacktrace: no debug info in Mach-O executable"

        - probably the same as https://github.com/ianlancetaylor/libbacktrace/issues/53

        ! winch segfaults, running gdb requires code-signing it

    ! excluded from build

- Tested inside RStudio

## Next steps

- file names in backtrace

    - Windows + Linux: easier in libbacktrace
    
    - macOS: requires debugging

- test C++ output, demangle

- get feedback from R core, especially Luke Tierney

- collection of IP addresses and function information for profiling


## History: jointprof part 1

- use gperftools to record in parallel

- requires parsing/processing alien data format: helper package

- works only on Linux, requires patch on Mac, overall very brittle

- hiatus, proffer package donated


## Further documentation links

- symbolic backtrace:

    - dwarf (http://kamalmarhubi.com/blog/2016/07/25/some-things-i-learned-about-libdwarf/), elf, ...

! related project in Nim: https://github.com/timotheecour/Nim/issues/49

## Related

! r-windows: submitted ccache pull request -- package and usage

    ! https://github.com/r-windows/rtools-packages/pull/147

    - pending: caching for R packages, https://github.com/r-windows/rtools-packages/issues/143

    - install or provide ag on msys2: https://packages.msys2.org/package/mingw-w64-x86_64-ag?repo=mingw64

! libbacktrace: implemented GitHub Actions, https://github.com/ianlancetaylor/libbacktrace/pull/51


! gcc has libbacktrace built in

    ! works, is included on Ubuntu, but not on Windows

    ! inclusion in Debian/Ubuntu seems to be a maintainer choice, https://bugs.gentoo.org/552098

    ! https://gcc.gnu.org/bugzilla/show_bug.cgi?id=66570

! three shells: msys2 for the system (incl. pacman), mingw64 and mingw32 for the build systems


! Accessing R_GlobalContext gives CRAN error


## Technical details

- sjlj vs. seh (why the difference between 32 and 64 bits?):

    - https://stackoverflow.com/q/15670169/946850

    - https://stat.ethz.ch/pipermail/r-sig-windows/2015q3/000020.html

    - http://mingw-w64.org/doku.php/contribute


## Trivia

! build gcc package in configured rtools, check that it can be installed

! double-check ccache in clean rtools, document usage for R packages

    - need to wait until it's synced with CRAN, or install manually
